---
title: "SARforecastDLM"
subtitle: "Code to generate sar_raw_data file"
output: html_document
date: "2025-01-13"
---

Objective: To generate the base data files for the SARforecastDLM package which include: 

- `sar_raw_data.rda`: includes all years of sar per method and index in one file -- used to model (`mod_mainpage_fct_forecast_model.R`) based on user selected years of input to generate `mod_mainpage_fct_forecast_compare_plot.R`


To update sar_raw_data, this script is setup up to call all new covariate data and then run through each SAR method to append into a single data file. Best to run in order-- will have to recode slighlty to update a single covariate/sar 

```{r load_libraries}
library(tidyverse)
library(here)
library(MARSS)
library(rerddap) #import data from ERDDAP
library(readxl) #import excel files--NCBI
library(lubridate)
```

# Covariate data

## Northern Copepod Biomass 
- Years: 1997 to current
```{r load_northern_copepod_biomass_data}

#Northern copepod biomass data
#NCBI<-read_csv(here("data-raw", "northern_copepod_biomass_newportOR.csv")) 

url<-"https://www.fisheries.noaa.gov/s3/2023-10/TempOxyNoSo-10232023.xlsx"

# Download the file to a temporary location
temp_file <- tempfile(fileext = ".xlsx")
download.file(url, temp_file, mode = "wb")

ncbi_raw<-read_excel(temp_file)


# Currently set to only pull April data but NOAA stoplight chart returns value for May - September. Future Research direction is attempting various month iterations.
df_ncbi<-ncbi_raw %>% 
  filter(Month == 4) %>% #select only april for now-- update with more research
  group_by(Year) %>% 
  summarise(value = mean(NORTHERN)) %>% 
  select("year" = Year, value) %>% 
  mutate(index = "NCBI")

ggplot(df_ncbi, aes(x = year, y = value)) +
  geom_line() +
  geom_point() +
  labs(title = "Northern Copepod Biomass",
       x = "Year",
       y = "Biomass (mg C m-3)[April]") +
  theme_minimal()

```

## Index of Coastal Prey Biomass (ICPB) 

- Years: 1998 to current
- Indice is of Winter (January, February, March) Ichthyoplankton 

- shared with JG via Elizabeth Daly - NOAA Affiliate see [Winter Ichthyoplankton via fisheries.noaa.gov](https://www.fisheries.noaa.gov/west-coast/science-data/2023-summary-ocean-ecosystem-indicators#winter-ichthyoplankton:~:text=at%20nearshore%20stations.-,Winter%20Ichthyoplankton,-The%202023%20winter). 

- Data in-house 1998 to 2019, emailed for more up to date data and directed to NOAA Stoplight Table for up-to-date data.  

- Since starts in 1998, will not include with SW SAR method, could look to older salmon prey biomass index (pre-ICPB). 

```{r load_ICPB_data}
#Northern copepod biomass data (in-house)
# df_ICPB<-read_csv(here("data-raw", "IndexOfCoastalPreyBiomass.csv")) 
 
#to pull directly from stoplight table 
library(data.table)
year <- 2024 #update most recent year
url <- paste0("https://www.fisheries.noaa.gov/s3/2024-12/stoplight-raw-data-",year,".csv")
raw_data<-fread(url, header = TRUE, check.names = FALSE)

df_icpb<-raw_data %>% 
  mutate(across(-1, as.numeric)) %>%
  pivot_longer(-1, names_to = "year", values_to = "value") %>% 
  filter(`Ecosystem Indicators` == "Nearshore Ichthyoplankton\nLog(mg C 1,000 m-3; Jan-Mar)") %>% 
  select(year, value) %>% 
  mutate(index = "ICPB") %>% 
  mutate(year = as.integer(year))

ggplot(df_icpb, aes(x = year, y = value)) +
  geom_line() +
  geom_point() +
  labs(title = "Nearshore Ichthyoplankton",
       x = "Year",
       y = "Biomass (mg C m-3)[April]") +
  theme_minimal()
```

## Coastal Upwelling Index (CUI)
 - Years: 1964 to current
```{r load_cui_data}
### ERDAPP, NOAA upwelling indice CUI (BAKUN 3 degree)
# Set the URL of the ERDDAP server
url <- "https://upwell.pfeg.noaa.gov/erddap/"

# Get information about the erdCUTI dataset
info <- info(datasetid = "erdUI45mo", url = url)
info # copy max range value from metadata


# Download the dataset
datcui <- griddap(info,
               time = c("1964-01-01T00:00:00Z", "2025-11-15T00:00:00Z"), #currently pulling all data to date (adjust as needed)
               latitude = c(45,45)
)

# format time and lat/long
df_cui_raw<-datcui$data %>% 
  mutate(time.day = lubridate::ymd_hms(time),
         year = year(time.day),
         month = month(time.day),
         day = day(time.day), 
         adjlong = longitude-360) #set to -180 to 180

## only include april CUI from ERDAPP data (for now)
df_cui<-df_cui_raw %>% 
  filter(month %in% c(4,5,6)) %>% 
         mutate(Month = month(month, label=TRUE)) %>% 
  select(year, Month, CUI=upwelling_index) %>% 
 #   rename("value" = upwelling_index)  %>%
  #mutate(index = "CUI")
    pivot_wider(names_from = Month, values_from=CUI)

df_cui_apr<- df_cui %>% select(CUI=Apr, year)
df_cui_mean<- df_cui %>% group_by(year) %>% summarise(mean_CUI=mean(c(Apr, May, Jun)))


ggplot(df_cui_raw %>% filter(month %in% c(4,5,6)) , aes(x= year, group=month, color=month))+
  geom_line( aes( y = upwelling_index)) +
  geom_line(aes(y = upwelling_index_anomaly), color = "red")
```


## CUTI
- Years: 1988 to current

```{r load_cuti_data}

#load CUTI
# Set the URL of the ERDDAP server
#url <- "https://upwell.pfeg.noaa.gov/erddap/"

# Get information about the erdCUTI dataset
info <- info(datasetid = "erdCUTImonthly", url = url)
info #copy max time range from metadata

# Download the dataset
datCUTI <- griddap(info,
               time = c("1988-01-15T00:00:00Z", "2025-03-15T00:00:00Z"), #currently pulling all data to date
               latitude = c(45,45)
)


df_cuti<-datCUTI$data %>% 
  mutate(time.day = lubridate::ymd_hms(time),
         year = year(time.day),
         month = month(time.day),
         day = day(time.day),
         Month = month(month, label=TRUE)) %>% 
  filter(month %in% c(4, 5, 6)) %>% 
  select(year, Month, CUTI)  %>% 
 # rename(value = "CUTI") %>% 
  #mutate(index = "CUTI") %>% 
    pivot_wider(names_from = Month, values_from=CUTI)

df_cuti_apr<- df_cuti %>% select(CUTI=Apr, year)
df_cuti_mean<- df_cuti %>% group_by(year) %>% summarise(mean_CUTI=mean(c(Apr, May, Jun)))

ggplot(df_cuti, aes(x = year, y = Apr)) +
  geom_line() +
  geom_point() +
  labs(title = "CUTI",
       x = "Year",
       y = "index") +
  theme_minimal()

```

```{r load_beuti_data}

#load BEUTI
# Set the URL of the ERDDAP server
#url <- "https://upwell.pfeg.noaa.gov/erddap/"

# Get information about the erdCUTI dataset
info <- info(datasetid = "erdBEUTImonthly", url = url)
info #copy max time range from metadata

# Download the dataset
datBEUTI <- griddap(info,
               time = c("1988-01-15T00:00:00Z", "2025-01-20T00:00:00Z"), #currently pulling all data to date
               latitude = c(45,45)
)


df_beuti<-datBEUTI$data %>% 
  mutate(time.day = lubridate::ymd_hms(time),
         year = year(time.day),
         month = month(time.day),
         day = day(time.day),
         Month = month(month, label=TRUE)) %>% 
  filter(month %in% c(4, 5, 6)) %>% 
  select(year, BEUTI, Month)  %>% 
  #rename(value = "BEUTI") %>% 
  #mutate(index = "BEUTI") %>% 
  pivot_wider(names_from = Month, values_from=BEUTI)

df_beuti_apr<- df_beuti %>% select(BEUTI=Apr, year)
df_beuti_mean<- df_beuti %>% group_by(year) %>% summarise(mean_BEUTI=mean(c(Apr, May, Jun)))

ggplot(df_beuti, aes(x = year, y = Apr)) +
  geom_line() +
  geom_point() +
  labs(title = "BEUTI",
       x = "Year",
       y = "index") +
  theme_minimal()

```

```{r load_cuti_beuti_daily_data}

#load CUTI
# Set the URL of the ERDDAP server
#url <- "https://upwell.pfeg.noaa.gov/erddap/"

# Get information about the erdCUTI dataset
info <- info(datasetid = "erdCUTIdaily", url = url)
info #copy max time range from metadata

# Download the dataset
datCUTIdaily <- griddap(info,
               time = c("1988-01-15T00:00:00Z", "2025-03-15T00:00:00Z"), #currently pulling all data to date
               latitude = c(45,45)
)

info <- info(datasetid = "erdBEUTIdaily", url = url)
info #copy max time range from metadata

# Download the dataset
datBEUTIdaily <- griddap(info,
               time = c("1988-01-15T00:00:00Z", "2025-01-20T00:00:00Z"), #currently pulling all data to date
               latitude = c(45,45)
)

df_cuti_daily<-datCUTIdaily$data %>% 
  mutate(time.day = lubridate::ymd_hms(time),
         year = year(time.day),
         month = month(time.day),
         day = day(time.day),
         Month = month(month, label=TRUE)) %>% 
  filter(month %in% c(4, 5, 6))

df_beuti_daily<-datBEUTIdaily$data %>% 
  mutate(time.day = lubridate::ymd_hms(time),
         year = year(time.day),
         month = month(time.day),
         day = day(time.day),
         Month = month(month, label=TRUE)) %>% 
  filter(month %in% c(4, 5, 6))
#Join BEUTI AND CUTI 
df.beuti.cuti<-df_beuti_daily %>% inner_join(df_cuti_daily, by = c("time.day", "day", "Month","month", "year"))

#Get monthly mean of mixed layer depth nitrate concentration
df_mldno3<- df.beuti.cuti %>% mutate(mldno3=BEUTI/CUTI) %>% group_by(month, year) %>% summarise(month.mean=mean(mldno3))

df_mldno3_apr<- df_mldno3 %>% filter(month==4) %>% rename(Apr.mldno3=month.mean)

df_mldno3_mean<- df_mldno3 %>% group_by(year) %>% summarise(mean_mldno3=mean(month.mean))

ggplot(df_mldno3 %>% filter(month==4), aes(x = year, y = month.mean)) +
  geom_line() +
  geom_point() +
  labs(title = "MLDNO3",
       x = "Year",
       y = "Nitrate") +
  theme_minimal()

```

```{r LWG Outflow_data}

o_url<- "https://www.cbr.washington.edu/dart/cs/php/rpt/mg.php?sc=1&mgconfig=river&outputFormat=csvSingle&year%5B%5D=2026&year%5B%5D=2025&year%5B%5D=2024&year%5B%5D=2023&year%5B%5D=2022&year%5B%5D=2021&year%5B%5D=2020&year%5B%5D=2019&year%5B%5D=2018&year%5B%5D=2017&year%5B%5D=2016&year%5B%5D=2015&year%5B%5D=2014&year%5B%5D=2013&year%5B%5D=2012&year%5B%5D=2011&year%5B%5D=2010&year%5B%5D=2009&year%5B%5D=2008&year%5B%5D=2007&year%5B%5D=2006&year%5B%5D=2005&year%5B%5D=2004&year%5B%5D=2003&year%5B%5D=2002&year%5B%5D=2001&year%5B%5D=2000&year%5B%5D=1999&year%5B%5D=1998&year%5B%5D=1997&year%5B%5D=1996&year%5B%5D=1995&year%5B%5D=1994&year%5B%5D=1993&year%5B%5D=1992&year%5B%5D=1991&year%5B%5D=1990&year%5B%5D=1989&year%5B%5D=1988&year%5B%5D=1987&year%5B%5D=1986&year%5B%5D=1985&year%5B%5D=1984&year%5B%5D=1983&year%5B%5D=1982&year%5B%5D=1981&year%5B%5D=1980&year%5B%5D=1979&year%5B%5D=1978&year%5B%5D=1977&year%5B%5D=1976&year%5B%5D=1975&year%5B%5D=1974&year%5B%5D=1973&year%5B%5D=1972&year%5B%5D=1971&year%5B%5D=1970&year%5B%5D=1969&year%5B%5D=1968&year%5B%5D=1967&year%5B%5D=1966&year%5B%5D=1965&year%5B%5D=1964&loc%5B%5D=LWG&data%5B%5D=Outflow&startdate=1%2F1&enddate=12%2F31&avgyear=0&consolidate=1&grid=1&y1min=0&y1max=&y2min=&y2max=&size=large"

Outflow<- read.csv(o_url) %>% filter(complete.cases(.)) %>% mutate(month_date=format(as.Date(paste0(mm.dd,'-',year), "%m-%d-%Y"))) %>% mutate(month=month(month_date), Year=year(month_date))

df_outflow<- Outflow %>% filter(month %in% c(4,5,6)) %>% group_by(Year) %>% summarise(mean.out=mean(value)) %>% rename(year=Year) 

ggplot(df_outflow, aes(x = year, y = mean.out)) +
  geom_line() +
  geom_point() +
  labs(title = "Outflow",
       x = "Year",
       y = "Outflow") +
  theme_minimal()

```
```{r Lower Goose Temperature_data}
#Temperature not recorded during first weeks of march
url_tLGO<- "https://www.cbr.washington.edu/dart/cs/php/rpt/mg.php?sc=1&mgconfig=river&outputFormat=csvSingle&year%5B%5D=2026&year%5B%5D=2025&year%5B%5D=2024&year%5B%5D=2023&year%5B%5D=2022&year%5B%5D=2021&year%5B%5D=2020&year%5B%5D=2019&year%5B%5D=2018&year%5B%5D=2017&year%5B%5D=2016&year%5B%5D=2015&year%5B%5D=2014&year%5B%5D=2013&year%5B%5D=2012&year%5B%5D=2011&year%5B%5D=2010&year%5B%5D=2009&year%5B%5D=2008&year%5B%5D=2007&year%5B%5D=2006&year%5B%5D=2005&year%5B%5D=2004&year%5B%5D=2003&year%5B%5D=2002&year%5B%5D=2001&year%5B%5D=2000&year%5B%5D=1999&year%5B%5D=1998&year%5B%5D=1997&year%5B%5D=1996&year%5B%5D=1995&year%5B%5D=1994&year%5B%5D=1993&year%5B%5D=1992&year%5B%5D=1991&year%5B%5D=1990&year%5B%5D=1989&year%5B%5D=1988&year%5B%5D=1987&year%5B%5D=1986&year%5B%5D=1985&year%5B%5D=1984&year%5B%5D=1983&year%5B%5D=1982&year%5B%5D=1981&year%5B%5D=1980&year%5B%5D=1979&year%5B%5D=1978&year%5B%5D=1977&year%5B%5D=1976&year%5B%5D=1975&loc%5B%5D=LGS&data%5B%5D=Temp+%28WQM%29&startdate=1%2F1&enddate=12%2F31&avgyear=0&consolidate=1&grid=1&y1min=0&y1max=&y2min=&y2max=&size=large&resolution=medium"

Temperature_LGO <- read.csv(url_tLGO) %>% filter(complete.cases(.)) %>% mutate(month_date=format(as.Date(paste0(mm.dd,'-',year), "%m-%d-%Y"))) %>% mutate(month=month(month_date), Year=year(month_date)) %>% filter(month %in% c(4,5,6))

#Get monthly means and filter out March because daily means not available for all of march
df_temp_LGO<- Temperature %>% group_by(Year) %>% summarise(mean.temp=mean(value)) %>% rename(year=Year)

ggplot(df_temp , aes(x = year, y = mean.temp)) +
  geom_line() +
  geom_point() +
  labs(title = "Temperature",
       x = "Year",
       y = "Degrees (C)") +
  theme_minimal()
```

```{r LWG Temperature_data}
#Temperature not recorded during first weeks of march
url_t<- "https://www.cbr.washington.edu/dart/cs/php/rpt/mg.php?sc=1&mgconfig=river&outputFormat=csvSingle&year%5B%5D=2026&year%5B%5D=2025&year%5B%5D=2024&year%5B%5D=2023&year%5B%5D=2022&year%5B%5D=2021&year%5B%5D=2020&year%5B%5D=2019&year%5B%5D=2018&year%5B%5D=2017&year%5B%5D=2016&year%5B%5D=2015&year%5B%5D=2014&year%5B%5D=2013&year%5B%5D=2012&year%5B%5D=2011&year%5B%5D=2010&year%5B%5D=2009&year%5B%5D=2008&year%5B%5D=2007&year%5B%5D=2006&year%5B%5D=2005&year%5B%5D=2004&year%5B%5D=2003&year%5B%5D=2002&year%5B%5D=2001&year%5B%5D=2000&year%5B%5D=1999&year%5B%5D=1998&year%5B%5D=1997&year%5B%5D=1996&year%5B%5D=1995&year%5B%5D=1994&year%5B%5D=1993&year%5B%5D=1992&year%5B%5D=1991&year%5B%5D=1990&year%5B%5D=1989&year%5B%5D=1988&year%5B%5D=1987&year%5B%5D=1986&year%5B%5D=1985&year%5B%5D=1984&year%5B%5D=1983&year%5B%5D=1982&year%5B%5D=1981&year%5B%5D=1980&year%5B%5D=1979&year%5B%5D=1978&year%5B%5D=1977&year%5B%5D=1976&year%5B%5D=1975&year%5B%5D=1974&year%5B%5D=1973&year%5B%5D=1972&year%5B%5D=1971&year%5B%5D=1970&year%5B%5D=1969&year%5B%5D=1968&year%5B%5D=1967&year%5B%5D=1966&year%5B%5D=1965&year%5B%5D=1964&loc%5B%5D=LWG&data%5B%5D=Temp+%28WQM%29&startdate=1%2F1&enddate=12%2F31&avgyear=0&consolidate=1&grid=1&y1min=0&y1max=&y2min=&y2max=&size=large"

Temperature <- read.csv(url_t) %>% filter(complete.cases(.)) %>% mutate(month_date=format(as.Date(paste0(mm.dd,'-',year), "%m-%d-%Y"))) %>% mutate(month=month(month_date), Year=year(month_date)) %>% filter(month %in% c(4,5,6))

LWG_LGO_temp<- full_join(Temperature, Temperature_LGO, by="month_date") %>% mutate(temp=ifelse(is.na(value.x), value.y, value.x), Year=ifelse(is.na(year.x), year.y, year.x))

#Get monthly means and filter out March because daily means not available for all of march
df_temp<- LWG_LGO_temp %>% group_by(Year) %>% summarise(mean.temp=mean(temp)) %>% ungroup()

saveRDS(df_temp, "C:/Users/krmacdon/Documents/CBR/DLM_Analysis_Git/data/Updated_temp_Jul26.rds")

ggplot(df_temp , aes(x = Year, y = mean.temp)) +
  geom_line() +
  geom_point() +
  labs(title = "Temperature",
       x = "Year",
       y = "Degrees (C)") +
  theme_minimal()
```


```{r PNI_data}
PNI <- read.csv(here( "data/pni_data.csv")) %>% select(Year, Annual.PNI) %>% rename(year=Year)
```

```{r SSTarc_data}
url<- "https://cbr.washington.edu/dart/cs/php/rpt/basin_scale_indices.php?sc=1&outputFormat=csv&indices=sstarc&syear=All&eyear=All"
dat_SST<- read.csv(url) %>% filter(complete.cases(.))

SST_prev <- dat_SST %>%
  group_by(Start.Month) %>% # Group by ID to compare within the same ID
  arrange(Year) %>% # Sort by year to ensure correct lag calculation
  mutate(Previous_Year_Value = lag(Value, n = 1), year=as.integer(Year)) %>%
  ungroup() %>% mutate(SST=ifelse(Start.Month==12, Previous_Year_Value, Value)) %>% select(-Year)

SST_pre1212<- SST_prev %>% filter(Start.Month %in% c(12, 1, 2)) %>% group_by(year) %>%  summarise(SST_pre1212=mean(SST))
SST_entry567<- SST_prev %>% filter(Start.Month %in% c(5, 6, 7)) %>% group_by(year) %>% summarise(SST_entry567=mean(SST))
SST_6mo<- SST_prev %>% filter(Start.Month %in% c(12, 1, 2, 3, 4, 5)) %>% group_by(year) %>% summarise(SST_6mo=mean(SST))
```
# SAR methods
All sar datasets have an additional blank logit.s year for ahead forecast in shiny app. 

```{r Spill}
url<- "https://www.cbr.washington.edu/dart/cs/php/rpt/mg.php?sc=1&mgconfig=river&outputFormat=csvSingle&year%5B%5D=2026&year%5B%5D=2025&year%5B%5D=2024&year%5B%5D=2023&year%5B%5D=2022&year%5B%5D=2021&year%5B%5D=2020&year%5B%5D=2019&year%5B%5D=2018&year%5B%5D=2017&year%5B%5D=2016&year%5B%5D=2015&year%5B%5D=2014&year%5B%5D=2013&year%5B%5D=2012&year%5B%5D=2011&year%5B%5D=2010&year%5B%5D=2009&year%5B%5D=2008&year%5B%5D=2007&year%5B%5D=2006&year%5B%5D=2005&year%5B%5D=2004&year%5B%5D=2003&year%5B%5D=2002&year%5B%5D=2001&year%5B%5D=2000&year%5B%5D=1999&year%5B%5D=1998&year%5B%5D=1997&year%5B%5D=1996&year%5B%5D=1995&year%5B%5D=1994&year%5B%5D=1993&year%5B%5D=1992&year%5B%5D=1991&year%5B%5D=1990&year%5B%5D=1989&year%5B%5D=1988&year%5B%5D=1987&year%5B%5D=1986&year%5B%5D=1985&year%5B%5D=1984&year%5B%5D=1983&year%5B%5D=1982&year%5B%5D=1981&year%5B%5D=1980&year%5B%5D=1979&year%5B%5D=1978&year%5B%5D=1977&year%5B%5D=1976&year%5B%5D=1975&loc%5B%5D=LWG&data%5B%5D=Spill+Percent&startdate=1%2F1&enddate=12%2F31&avgyear=0&consolidate=1&grid=1&y1min=0&y1max=&y2min=&y2max=&size=large&resolution=medium"

dat_Spill<- read.csv(url) %>% filter(complete.cases(.)) %>% mutate(month=as.integer(str_sub(mm.dd, start=1, end=1)))

mean_spill<- dat_Spill %>% filter(month %in% c(4, 5, 6)) %>% group_by(year) %>% summarise(per_spill=mean(value))

saveRDS(mean_spill, "mean_spill.rds")
```

## Scheuerell and Williams (2005) 
- Years: 1964 to 2005
- Reaches: LGR_LGA (ish)
- Loaded data via  `SalmonSurvCUI`, `MARSS` package

```{r load_MARSS_sar_data}
## load the data
data(SalmonSurvCUI, package = "MARSS")


sar_raw_data <- SalmonSurvCUI %>% 
  rename(SW_logit_phi=logit.s) %>% 
  mutate(year = as.numeric(year)) %>% 
  select(-CUI.apr)


#join with CUTI data only (others do not have enough older years overlap), includes NA for 2006
#sar_raw_data_CUI_CUTI<-sar_raw_data %>% 
#  left_join(df_cuti, by = "year") %>% 
#  bind_rows(sar_raw_data) %>% 
#  mutate( reach = "LGR_LGA")

```

## CBR DART SAR data
- Reaches: LGR_LGA, BON_BOA
To generate CBR sar data, used https://www.cbr.washington.edu/dart/query/pit_sar_esu query tool with selections: 
  - Lower Granite (All) to Lower Granite Adult or Bonneville (All) to Bonneville Adult
  - Chinook Snake R Spring/Summer ESU
  - Rear Type : Wild
  - Adult Detections : Exclude 0,1-Year Adult Detections
  - Set Annual Summary Table Group Type: by Release Site
  - download Pooled by Year SAR csv file link
  - downloaded, saved in `data-raw/SAR` folder, and renamed `LGRLGA_allpassage_wild_spsu_chinook.csv` 
  - per [issue #4](https://github.com/Columbia-Basin-Research-CBR/SARforecastDLM/issues/4) will limit analysis of DART data to 2002 to most recent outmigration year (2021); reconsider if using different variance matrix- check it still will not converge if included.
```{r load_LGRLGA_dart_sar_data}
#generated new file 1/13/25 with updated years of sar (2000 to 2021) changed values from previous `LGRtoLGA_allpass_W_spsu_ch.csv` file (added 2000-2001; updated 2021 sar values)

load(here( "data/LGRLGA_allpassage_wild_spsu_chinook.rda"))
df.cbr.raw<- LGRLGA_allpassage_wild_spsu_chinook %>% 
  drop_na() %>% 
  select(year, meanSAR) %>% #drop CBR notes and NA values
  mutate(year = as.integer(year))

# must divide meanSAR /100 to get 0 to 1 and then apply logit/transform since DART sar calculated as adultCount/juvCount*100
df.cbr<-df.cbr.raw %>% 
  mutate(SAR_logit_phi = qlogis(meanSAR/100)) %>% #
  select(-meanSAR)
 # mutate(sar.method = "DART",
#         reach = "LGR_LGA") %>% 
#  add_row(year = max(.$year)+1, logit.s = NA, sar.method = "DART", reach = "LGR_LGA") #add next year with NA


#CBR SAR + April data
df.cbr.LGRLGA.apr<-df.cbr %>% inner_join(df_cui_apr, by = "year")  %>% inner_join(df_cuti_apr, by = "year") %>% inner_join(df_beuti_apr, by = "year")  %>% inner_join(df_mldno3_apr, by = "year")  %>% inner_join(df_outflow, by = "year")  %>% inner_join(df_temp, by = "year")  %>% inner_join(PNI, by = "year")  %>% inner_join(df_cui_mean, by = "year") %>% inner_join(df_cuti_mean, by = "year")  %>% inner_join(df_beuti_mean, by = "year")  %>% inner_join(df_mldno3_mean, by = "year")

#Save CBR SAR April data
saveRDS(df.cbr.LGRLGA.apr, "df_cbr_LGRLGA_apr.rds")
```


```{r load_BONBOA_dart_sar_data}
#generated new file 1/13/25 with updated years of sar (2000 to 2021) changed values from previous `LGRtoLGA_allpass_W_spsu_ch.csv` file (added 2000-2001; updated 2021 sar values)

load(here("data", "BONBOA_allpassage_wild_spsu_chinook.rda"))
df.cbr.raw<-BONBOA_allpassage_wild_spsu_chinook %>% 
  drop_na() %>% 
  select(year, meanSAR) %>% #drop CBR notes and NA values
  mutate(year = as.integer(year))

# must divide meanSAR /100 to get 0 to 1 and then apply logit/transform since DART sar calculated as adultCount/juvCount*100
df.cbr<-df.cbr.raw %>% 
  mutate(logit.s = qlogis(meanSAR/100)) %>% #
  select(-meanSAR) %>%
  mutate(sar.method = "DART",
         reach = "BON_BOA") %>% 
  add_row(year = max(.$year)+1, logit.s = NA, sar.method = "DART", reach = "BON_BOA") #add next year with NA


#CUI
df.cbr.cui<-df.cbr %>% inner_join(df_cui, by = "year")

#CUTI
df.cbr.cuti<-df.cbr %>% inner_join(df_cuti, by = "year")

#ICPB
df.cbr.icpb<-df.cbr %>% inner_join(df_icpb, by = "year") 
#NCBI
df.cbr.ncbi<-df.cbr %>% inner_join(df_ncbi, by = "year") 


#join all indices
df.cbr<-rbind(df.cbr.cui, df.cbr.cuti, df.cbr.icpb, df.cbr.ncbi) 


#add to sar_raw_updated with DART sar method with CUI,CUTI, NCBI
sar_raw_data<-rbind(sar_raw_data, df.cbr)
```

## CJS estimated survival
- Years: 1994 to 2021
- Reaches: LGR_LGA, BON_BOA
- This has been updated with JG generated data-- includes which covariates? Just transport?
- To update this with freshwater covariates (transport,temperature, etc)

```{r load_LGRLGA_cjs_sar_data}

df_cjs_raw<-read_csv(here( "data/lgr.lga.surv.yr.1994.2021_250309.csv")) 
df_cjs_raw <- df_cjs_raw[-1]


df.cjs <- df_cjs_raw %>% 
  mutate(cjs_logit_phi = qlogis(LGR_LGA_SAR)) %>% 
  select(-LGR_LGA_SAR) 
 # mutate(sar.method = "CJS", 
#         reach = "LGR_LGA") %>% 
#  add_row(year = max(.$year)+1, logit.s = NA, sar.method = "CJS", reach = "LGR_LGA") #add 2020 with NA

df_SAR_LGRLGA_MV<-sar_raw_data %>% full_join(df.cbr, by="year") %>%  full_join(df.cjs, by="year") %>% left_join(df_cui_apr, by = "year")  %>% left_join(df_cuti_apr, by = "year") %>% left_join(df_beuti_apr, by = "year")  %>% left_join(df_mldno3_apr, by = "year")  %>% left_join(df_outflow, by = "year")  %>% left_join(df_temp, by = "year")  %>% left_join(PNI, by = "year") %>% left_join(df_cui_mean, by = "year") %>% left_join(df_cuti_mean, by = "year")  %>% left_join(df_beuti_mean, by = "year")  %>% left_join(df_mldno3_mean, by = "year") %>% left_join(SST_pre1212, by = "year") %>% left_join(SST_entry567, by = "year") %>% left_join(SST_6mo, by = "year") %>% 
  select(-c(month))

#Save LGRLGA CJS data
saveRDS(df_SAR_LGRLGA_MV, "df_SAR_LGRLGA_MV.rds")
```

```{r load_BONBOA_cjs_sar_data}

df_cjs_raw<-read_csv(here("bon.boa.surv.yr.1994.2021_250309.csv")) 
df_cjs_raw <- df_cjs_raw[-1]




df.cjs <- df_cjs_raw %>% 
  mutate(logit.s = qlogis(BON_BOA_SAR)) %>% 
  select(-BON_BOA_SAR) %>% 
  mutate(sar.method = "CJS", 
         reach = "BON_BOA") %>% 
  add_row(year = max(.$year)+1, logit.s = NA, sar.method = "CJS", reach = "BON_BOA") #add 2020 with NA


#CUI
df.cjs.cui<- df.cjs %>% inner_join(df_cui, by = "year") 

saveRDS(df.cjs.cui, "df_cjs_cui.rds")
#CUTI
df.cjs.cuti<-df.cjs %>% inner_join(df_cuti, by = "year")

saveRDS(df.cjs.cuti, "df_cjs_cuti.rds")
#BEUTI
df.cjs.beuti<-df.cjs %>% inner_join(df_beuti, by = "year")
saveRDS(df.cjs.beuti, "df_cjs_beuti.rds")

#ICPB
df.cjs.icpb<-df.cjs %>% inner_join(df_icpb, by = "year")
#NCBI
df.cjs.ncbi<-df.cjs %>% inner_join(df_ncbi, by = "year") 


#join all indices
df.cjs<-rbind(df.cjs.cui, df.cjs.cuti, df.cjs.icpb, df.cjs.ncbi) 

#add to sar_raw_data
sar_raw_data<-rbind(sar_raw_data, df.cjs)
```

# save (update) sar_raw_data
As of 6/20/24 adding NCBI back into DLM as place holder for ICPB data. NCBI is publicly available whereas ICPB is not. Updated 1/13/25, per email exchange with E. Daly, ICPB data is publically available through NOAA Stoplight Chart. Adding back into option list with NCBI.  

```{r check_save_sar_raw_data}
#check each sar method has correct years and indices
sar_raw_data%>% 
  group_by(sar.method, index) %>% 
  summarise(n_distinct(year))


# plot sar on logit scale
sar_raw_data %>% 
  filter(index == "CUI") %>% 
  ggplot(aes(x = year)) +
  geom_line(aes(y = logit.s, color = sar.method))+
  geom_point(aes(y = logit.s, color = sar.method), shape = 21) +
  labs(title = "Observed SAR by method",
       x = "Year",
       y = "SAR (logs-odd)") +
  theme_minimal()

#plot inverse logit to check
sar_raw_data %>% 
  mutate(sar = plogis(logit.s)*100) %>% 
  filter(index == "CUI") %>% 
    ggplot(aes(x = year)) +
  geom_line(aes(y = sar, color = sar.method))+
  geom_point(aes(y = sar, color = sar.method), shape = 21) +
  labs(title = "Observed SAR by method",
       x = "Year",
       y = "SAR (%)") +
  theme_minimal()


#inverse matches original data --yes
df_marss<-SalmonSurvCUI %>% 
  mutate(sar = plogis(logit.s))
ggplot() +
  geom_line(data = df_marss, aes(x = year, y = sar*100), color = "red") + #, color = "Scheuerell and Williams (2005)")) +
  geom_line(data = df_cjs_raw, aes(x = year, y = LGR_LGA_SAR), color = "blue") + #, color = "CJS")) +
  geom_line(data = df.cbr.raw, aes(x = year, y = meanSAR), color = "green") + #, color = "DART")) +
    labs(title = "Observed SAR by method",
       x = "Year",
       y = "SAR (%)") +
  theme_minimal()
```


```{r save_sar_raw_data}
#save updated sar_raw_data
usethis::use_data(sar_raw_data, overwrite = TRUE)
```

